Kindly Bent to Free Us

### 2019-06-27 kindly/motivation.tex
Make match parameterized over its modality.
My brain is too warm.

We can then use these primitives to implement a simple addition service.
We do not need any extra linearity annotations in the code: the fact
that the \texttt{st} type is declared as linear is sufficient.
In the case below, \lstinline{add_service} is inferred to be of type
\lstinline{(int ! int ! int ? end) st -> unit}

\begin{lstlisting}
let add_service ep =
  let x, ep = receive ep in
  let y, ep = receive ep in
  let ep = send (x + y) ep in
  close ep
# add_service : (int ! int ! int ? end) st -> unit
\end{lstlisting}

The following code then acts as a client, by sending two messages
and receiving the result.
This code is polymorphic in both argument and return types and could thus
be used to any binary operator.
%
\begin{lstlisting}
let op_client ep x y =
  let ep = send x ep in
  let ep = send y ep in
  let result, ep = receive ep in
  close ep;
  result
# op_client : ('a_1 ? 'a_2 ? 'b ! end) st -> 'a_1 -{lin}> 'a_2 -{lin}> 'b
\end{lstlisting}
% The actual answer from the typechecker looks like this:
% # ('m2:^k2), ('m1:^k1).
% # (^k1 < ^k) & (lin < ^k) =>
% # ('m1 ?'m2 ? 'result ! end) st -> 'm1 -{lin}> 'm2 -{^k}> 'result

In addition, this function can be partially
applied to a channel. Since the closure returned by such a partial
application captures the channel, it can only be used once.
This is reflected by the arrow, \lstinline/-{lin}>/, which is the
type of functions that can only be used once.
The general form of arrows types in \lang is \lstinline/-{k}>/, where
\lstinline/kk/ is an arbitrary kind that describes the behavior of
underlying closure.
For convenience, we shorten \lstinline/-{un}>/ as \lstinline/->/.
Note again that the linearity on the arrows was inferred without
any annotation from the user. In fact, the only difference between the code
presented so far and the examples by
\citet{DBLP:journals/jfp/Padovani17}
is the kind annotation on the type definition of \lstinline/st/.

@@

In the last few years, new persistent data-structures
have been introduced which allow transient mutations, but discourage
non-linear uses either through degraded performances
\cite{DBLP:conf/ml/ConchonF07} or
dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
In particular, persistent Hash-Array-Mapped-Trie has been used with similar
API in numerous non-pure functional languages (OCaml, Clojure, \dots).
Affine types allow to formalize the performance contract between the programmer
and the library, while borrows avoid the need to thread state explicitly,
as usually required by an immutable API.
%
In this section, we show how linearity and borrowing allows us to define
a safe API for persistent arrays that support both immutable and mutable features,
and how it can be used to implement a backtracking sudoku solver.

@@

Thanks to this API, modifications that correspond to backtracking points
use \lstinline/set/, which allows arbitrary uses of the previous version
of the array, while other modifications uses \lstinline/set_mut/, which
forbids uses of the previous version.
This can be seen in the sudoku solver implementation in \cref{ex:sudoku}.
The board is represented as a matrix of cells (\cref{line:boardtype}).
We assume the existence of a \lstinline/Matrix/ module which uses the same API
as \lstinline/HybArray/ but with two indices.
Each cell is an integer
set that represent admissible solutions so far. The sudoku solver will
iterate on the cells and try each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we use an immutable modification
to create a backtracking point (\cref{line:immutset}) and propagate
the changes with the \lstinline/propagate/ function. The \lstinline/propagate/
function does not need to preserve the previous
version of the board, and thus uses direct mutation.
The implementation of \lstinline/propagate/ is split into three parts
for lines, columns, and square. For concision, only \lstinline/propagate_lines/ is
presented here (\cref{line:propline}).
The typing ensures that all the mutations do not compromise
backtracking, since sharing is only allowed through
the immutable set functions, which is used for backtracking points.

@@

\subsection{Type abstraction and linearity violation}

>>>>

We can then use these primitives to implement a simple addition service.
Both client and server code are shown in
\cref{fig:sessiontype}.
We do not need any extra linearity annotations in the code: the fact
that the \texttt{st} type is declared as linear is sufficient.
The server, \lstinline{add_service}, is inferred to be of type
\lstinline{(int ! int ! int ? end) st -> unit}.
The client, \lstinline/op_client/, operates by sending two messages
and receiving the result.
This code is polymorphic in both argument and return types and could thus
be used to any binary operator.
%
\begin{figure}[!h]
  \begin{subfigure}[t]{.5\linewidth}
\begin{lstlisting}
let add_service ep =
  let x, ep = receive ep in
  let y, ep = receive ep in
  let ep = send (x + y) ep in
  close ep
# add_service :
  (int ! int ! int ? end) st -> unit
\end{lstlisting}
    \caption{Addition server}
  \end{subfigure}~
  \begin{subfigure}[t]{.5\linewidth}
\begin{lstlisting}
let op_client ep x y =
  let ep = send x ep in
  let ep = send y ep in
  let result, ep = receive ep in
  close ep;
  result
# op_client :
  ('a_1 ? 'a_2 ? 'b ! end) st -> 'a_1 -{lin}> 'a_2 -{lin}> 'b
\end{lstlisting}
% The actual answer from the typechecker looks like this:
% # ('m2:^k2), ('m1:^k1).
% # (^k1 < ^k) & (lin < ^k) =>
% # ('m1 ?'m2 ? 'result ! end) st -> 'm1 -{lin}> 'm2 -{^k}> 'result
    \caption{Binary operators client}
  \end{subfigure}
  \caption{Corresponding session type programs in \lang}
  \label{fig:sessiontype}
\end{figure}

In addition, the \lstinline/op_client/ function can be partially
applied to a channel. Since the closure returned by such a partial
application captures the channel, it can only be used once.
This is reflected by the arrow, \lstinline/-{lin}>/, which is the
type of functions that can only be used once.
The general form of arrows types in \lang is \lstinline/-{k}>/, where
\lstinline/kk/ is an arbitrary kind that describes the behavior of
underlying closure.
For convenience, we shorten \lstinline/-{un}>/ as \lstinline/->/.
Note again that the linearity on the arrows was inferred without
any annotation from the user. In fact, the only difference between the code
presented so far and the examples by
\citet{DBLP:journals/jfp/Padovani17}
is the kind annotation on the type definition of \lstinline/st/.

@@

In the last few years, new persistent data-structures
have been introduced which allow transient mutations, but discourage
non-linear uses either through degraded performances
\cite{DBLP:conf/ml/ConchonF07} or
dynamic and static checks \cite{DBLP:journals/pacmpl/Puente17}.
In particular, persistent Hash-Array-Mapped-Trie has been used with similar
API in numerous non-pure functional languages (OCaml, Clojure, \dots).
Affine types allow to formalize the performance contract between the programmer
and the library, while borrows avoid the need to thread state explicitly,
as usually required by an immutable API.
%
In this section, we show how \lang allows us to define
a safe API for persistent arrays that support both immutable and mutable features,
and how it can be used to implement a simple backtracking sudoku solver.

@@

Thanks to this API, modifications that correspond to backtracking points
use \lstinline/set/, which allows arbitrary uses of the previous version
of the array, while other modifications uses \lstinline/set_mut/, which
forbids uses of the previous version.
This can be seen in the sudoku solver implementation in \cref{ex:sudoku}.
The board is represented as a matrix of cells (\cref{line:boardtype}).
We assume the existence of a \lstinline/Matrix/ module which uses the same API
as \lstinline/HybArray/ but with two indices.
Each cell is an integer
set that represent admissible solutions so far. The sudoku solver will
iterate on the cells and try each possible solution (\cref{line:try_solution}).
When a value is picked for the current cell, we use an immutable modification
to create a backtracking point (\cref{line:immutset}) and propagate
the changes with the \lstinline/propagate/ function. The \lstinline/propagate/
function does not need to preserve the previous
version of the board, and thus uses direct mutation.
The implementation of \lstinline/propagate/ is split into three parts
for lines, columns, and square. For concision, only \lstinline/propagate_lines/ is
presented here (\cref{line:propline}).
The typing ensures that all the mutations do not compromise
backtracking, since sharing is only allowed through
the immutable set functions, which is used for backtracking points.
In particular, the \lstinline/try_solution/ function can be used in a loop
precisely because the \lstinline/set/ function only requires an
\emph{immutable} borrow of \lstinline/new_g/.
This is allowed, since immutable borrows are unrestricted.

@@

\subsection{Type abstraction and linearity violations}
